### 2022-09-12 AISTATS2023/paper.tex

\paragraph{Notation.}

We refer to random variables (rv) using uppercase \( X \), whereas to
simple variables with plain lowercase \( x \). Bold denotes a vector;
\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
interest \(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} PDP formulates the feature effect
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects, which are defined as the
change on the output, i.e.
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDP integrates
over unrealistic instances due to the use of the marginal distribution
\( \mathbf{X}_c\). MPs resolve this issue using the conditional
distribution \( \math{C}_2|\mathcal{X}_1 \) but suffer from
computing combined effects. ALE plots resolve, both they have two
drawbacks TODO add more info.

TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.

\paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
feature effect methods to inform about the heterogeneous
effects. Elaborate. Interpretation of the heterogeneous effects behind
the global effect is available only for PDP, with three different
approaches; (a) ICE and d-ICE plots provide a visual understanding of
the heterogeneous effects. (b) grouping of ICE in homogeneous
clusters, for spliting the input space into subspace(s) with
homogeneous effects (c) Feature Interaction strength indexes, like
H-statistic, provide a value indicating how much a feature interacts
with the others (not the type of interaction). There is no method for
quantifying the heterogeneous effects, based on ALE. Therefore, no
method to exploit the advantages of ALE while, on the same time,
informing about the heterogeneous effects.

\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
axis into intervals, allocating the instances of the training set in
the intervals and compute a single (constant) effect in each
interval. With DALE extension, bin spliting is decoupled from instant
effect estimation. With our extension for measuring the heterogeneous
effects, we transfrom interval spliting from a step to a clustering
problem with a meaningful objective to minimise. We provide a thorough
analysis, where we show that our objective has a consistent
meaning. It can be split in two parts; the first part is the
unavoidable uncertainty due to the natural characteristics of the
experiment, i.e.,~the data generating distribution and the black-box
function. The second part is an added uncertainty due to the
limited-samples estimation, that enforces to create groups with
constant main effect. We opt for minimizing the objective, i.e. sum of
the two uncertainties, that given that the first uncertainty is
independent of the bin spliting, therefore we want to minimize the
added uncertainty. To conclude, we transform the axis-spliting into an
unsupervised clustering problem with a principled objective. We a
computationally-grounded solution that works out-of-the-box, relaxing
the user from providing a hyperparameter without any indication which
one is the correct. This step can be used independently of whether the
user wants to explore the heterogeneous effects or not.

\section{THE ... METHOD}

\subsection{ALE with Uncertainty Quantification}

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are weigthed by the conditional distribution
\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the uncertainty of the local
effects at \(\mathcal{H}(x_s)\) as the variance of the local
explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In Section
(TODO), we propose appropriate visualizations for easier
interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the feature effect
at \(x_s\) is the accumulation of the averaged local effects from
\(x_{min}\) until \(x_s\), as show in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until the point \(x_s\), as the
integral of the variances of local effects:

@@

The accumulated uncertainty is not a directly interpretable
quantity. It only helps us define a sensible objective for the
interval spliting step, as we discuss in Section TODO: adde ref.

\subsection{Uncertainty Quantification and Estimation at an Interval}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point \(x_s\) is impossible, because the probability of observing a
sample inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). Therefore, we are obliged to split the axis of \(x_s\)
into a sequence of non-overlaping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:

@@

\noindent
Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
is defined as:

\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}

@@

The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). (TODOs: Check what happens
otherwise).

\subsection{Bin Spliting as a Clustering Problem}

ALE, Eq.(\ref{eq:ALE_accumulated_mean}), is estimated by spliting the
axis \(x_s\) into a sequence of non-overlaping bins (TODO: add some
discussion for ALE, DALE):

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
\end{equation}

and

\begin{equation}
  \label{eq:ALE_accumulated_var_est}
  \hat{f}^{\mathtt{ALE}}_{\sigma^2}(x_s) = \sum_{k=1}^{k_x} \hat{\sigma}^2(z_{k-1}, z_k)
\end{equation}

We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). Both
methods face the limitation that the partitioning into non-overlaping
intervals is done blindly. The user pass the total number of bins
\(K\) as a hyperparameter, the bins are defined with equal-size
spliting, and the training instances are allocated accordingly. This
approach is vulnerable to non-robust estimations. The mean effect is
often poorly approximated from a very small number of samples and the
mean effect of empty bins is interpolated from their
neighboors. Furthermore, in our case, we need sufficient sample
populations for estimating the variance of the approximation, apart
from the mean effect.

\subsubsection{Methodology}

For overcoming this limitations, we reformulate the partitioning as a
clustering of the training instances into a sequence variable-size
intervals. The objective of the clustering problem is inspired

ALE requires splittingthe estimation of the

In this section, we introduce a framework

\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
at an interval \([z_1, z_2)\) is the accumulation of the all variances
plus the accumulation of squared residuals inside the interval:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is at the Appendix. Theorem 1 decouples the accumulated
variance at an interval, the only quantity we can estimate, into two
terms. The first term \(\int_{z_1}^{z_2} \sigma^2(z) \partial z\),
quantifies the uncertainty due to the natural characteristics of the
experiment and the second term adds extra uncertainty due to the
limited resolution.

\paragraph{Uncertainty of the global effect.}

Eq.~\eqref{eq:variance_estimation} gives an approximation of the
uncertainty of the bin effect.
The uncertainty of the global effect is
simply the sum of the uncertainties in the bin effects.

\paragraph{Minimizing the uncertainty}

@@

\subsection{Visualization of ALE with Uncertainty}

@@

\section*{Appendix}

\subsection{Proof for variance of the bin}

>>>>

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), whereas to simple variables with plain lowercase
\( x \). Bold denotes a vector; \( \xb \) for simple variables or
\(\mathbf{X}\) for rvs. Often, we partition the input vector
\(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} As stated at the Introduction, the
three well-known feature effect methods are: PDPs, MPs and ALE. PDPs
formulate the feature effect of the \(s\)-th attribute as an
expectation over the marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects, which are defined as the
change on the output, i.e.
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
corelated input features. In these cases, PDPs integrate over
unrealistic instances, due to the use of the marginal distribution
\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
impute the combined effect of sets of features to a single
feature. ALE manages to resolve both issues, and is therefore the only
trustable method in cases of correlated features.

\paragraph{Quantify the Heterogeneous Effects.}

Feature effect methods answer the question \textit{what happens
  (effect) to the output, if I increase/decrease the value of a
  specific feature}. Having answered the question above, it comes
naturally to also wonder \textit{how certain we are about the change
  on the output}. For this reason, a lot of interest is given lately
for quantifying the level of uncertainty, along with the expected
effect. The level of uncertainty is mostly quantified by measuring the
existence of heterogeneous effects, i.e. whether there are local
explanations that deviate from the expected global effect. ICE and
d-ICE plots provide a visual understanding of the heterogeneous
effects on top of PDPs. Another approach targets on grouping the
heterogeneous effects, e.g., allocating ICE plots in homogeneous
clusters, by dividing the input space. Some other approaches, like
H-Statistic, Greenwel, move a step behind and try to quantify the
level of interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches face two pathogenies; They
either do not quantify the uncertainty of the feature effect directly
or they are based on PDPs, and, therefore, they are subject to the
failure modes of PDPs in cases of correlated features. To the best of
our knowledge, no method so far targets on quantify the uncertainty of
the feature effect as it is modelled by ALE.

\paragraph{Cluster Instances with homogeneous effects.}

In real ML scenarios, the expected feature effect and the uncertainty
are estimated from the limited instances of the training set. ALE
approximation requires an additional step, where the axis of the
\(s\)-th feature is split into a sequence of non-overlaping bins and a
single effect (expectation and uncertainty) is computed from the
population of instances that lie inside each
bin. \citep{apley2020visualizing} proposed estimating the local
effects in each bin by evaluating the black box-funtion at the bin
limits:

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}

We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). In
contrast, (cite) proposed the Differential ALE (DALE) estimation for
quantifying the local effects on the training-set instances, instead
of the bin limits:

\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, it allows the recomputation
of the accumulated effect with different bin-splitting in near-zero
computational overhead. However, none of the approximations above
deals with the crucial problem of the optimal bin-spliting. They
partition the axis in \(K\) equally-sized bins without considering the
properties of the underlying data, which can lead to erroneous
approximations.

Instead, we propose treating the axis-spliting step as an unsupervised
clustering problem. The objective of the clustering problem should
fulfill in the best way to contradictory ojectives. First, secure
robust estimations of the expected effect and the uncertainty inside
each bin given the limited instances of the training set and, second,
create bins with as homogeneous local effects as possible, for not
losing fine-grain resolution feature effects due to wide bins.

\section{THE NAME METHOD}
\label{sec:NAME-method}

The NAME method extends the traditional ALE method with a component
for uncertainty quantification and improves the ALE estimation by
automatically discovering the optimal bin-splitting. In
Section~\ref{sec:NAME-definition} we define the component for the
uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
NAME from the limited samples of the training set and we make an
important proof about the aggregated variance defined over a bin. In
Section~\ref{sec:interval-spliting}, we define and solve the problem
of optimal bin-splitting. Finally, in Section~\ref{sec:visualization},
we illustrate the appropriate visualization of NAME for facilitating
its interpretation by a non-expert and we discuss important aspects of
the method

\subsection{Uncertainty Quantification}
\label{sec:NAME-definition}

ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are, then, weigthed by the conditional
distribution \(p(\xc|x_s)\) and are averaged, to produce the
summarized effect at \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The feature effect at \(x_s\) is the accumulation of the averaged
local effects from \(x_{s, min}\) until \(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). As described at the Introduction, limiting the explanation to the
exepected value level does not shed light to possible heterogeneous
effects behind the averaged explanation. Therefore, we model the
uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
feature effect at \(x_s\) is the accumulation of the averaged local
effects from \(x_{min}\) until \(x_s\), as shown in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until \(x_s\), as the integral of
the variances of the local effects:

@@

The accumulated uncertainty is not a directly interpretable
quantity. It only helps us define a sensible objective for the
interval spliting step, as we discuss in
Section~\ref{sec:interval-spliting}.

\subsection{Interval-Based Estimation}
\label{sec:interval-based-estimation}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
on the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point is impossible, because the probability of observing a sample
inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
into a sequence of non-overlaping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:

@@

\noindent
Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
defined as:

\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}
\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
at an interval \([z_1, z_2)\) is the accumulation of the all variances
plus the accumulation of squared residuals inside the interval:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix. Theorem 1 decouples the accumulated
variance at an interval into two terms. The first term
\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\), quantifies the aggregated
uncertainty due to the natural characteristics of the experiment
\(\mathcal{H}\) and the second term adds extra nuisance uncertainty
due to the limited resolution \(\mathcal{H}_n\). In other words,
enforcing the computation of a single effect for all points in
\([z_1, z_2)]\), burdens the estimation inside the interval with a
nuisance uncertainty of:

\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \int_{z_1}^{z_2} \mathcal{H}(z) + \mathcal{H}_n(z) \partial z
\end{equation}

@@

The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). Elaborate.

\subsection{Interval Spliting As A Clustering Problem}
\label{sec:interval-spliting}

We formulate the axis-splitting as an unsupervised clustering
problem. We search for the optimal bin splitting, i.e., the number and
size of consecutive non-overlapping intervals that minimizes the
accumulated variance. The optimization problem is defined as follows:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
\min_{ \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \\
\textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
\end{aligned}
\end{equation}

The objective of the optimization problem is the sum of the
accumulated variances in each bin, estimated by the instances of the
training set. As we prove in Theorem 1,
\(\mathcal{L} = \sum_{k=1}^K \mathcal{H}_{bin}(z_{k-1}, z_k)\) is the
sum of the uncertainty of the local explanations \(\mathcal{H}\) due
to heterogeneous effects and the added nuisance uncertainty due to
bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
affected by interval splitting, optimizing \(\mathcal{L}\) equals to
finding the sequence of non-overlaping bins that add the minimum
nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
each bin is populated with enough samples for a robust estimation,
which is set by the parameter \(N\). In other words, the user makes
the following proposal; \textit{find the bin sequence that adds the
  minimum nuisance uncertainty, given that I want at least \(N\)
  points per bin}.

\subsubsection{Algorithm For Solving The Clustering Problem}
\label{sec:dynamic-programing}

@@

\subsection{Visualization of NAME Method and Discussion}
\label{sec:visualization}

\begin{itemize}
\item Discuss about the meaning of ALE, to find intervals with some effect
\end{itemize}

@@
